---
title: "Code 1: css-normalization for 16S data"
author: "Bryan J. Rodriguez-Colon"
date: "Last udpated: 2025-09-10"
output: html_document
---


***Disclaimer: This code was designed by Bryan Rodriguez-Colon with the help of chatGPT-4o & chatGPT-5.***

This R Markdown performs **cumulative sum scaling (CSS) normalization** on 16S amplicon sequencing count data using the `metagenomeSeq` package. Input is a **BIOM file** with counts and taxonomy. The script includes the following steps:

1. Load the BIOM file into an `MRexperiment` object.
2. Filter out non-Bacteria/Archaea sequences.
3. Remove samples with zero counts to avoid normalization errors.
4. Apply CSS normalization.
5. Reintroduce empty samples as zero-filled columns for consistency.
6. Save the normalized data back as a BIOM file for downstream use.

CSS provides a principled alternative to rarefaction for normalizing sparse amplicon counts (Paulson et al., 2013).




```{r setup, include=T}
#Set up.
knitr::opts_chunk$set(echo = TRUE)

#Working Directory
setwd("/Users/bryanrodriguez/Library/CloudStorage/OneDrive-UniversityofKansas/Bryan-MS_PhD/PHD-PROJECT/PuertoRico/CHAPTER2/16S/Processing/Geobio_Guanica_SciRep_Codes/R_Codes")

#Clean Environment
rm(list = ls())

```


```{r Load packages, include=T, message=F}

# Load necessary packages for normalization and data handling
library(metagenomeSeq)  # For CSS normalization and MRexperiment objects
library(biomformat)     # For reading/writing BIOM files
library(Biobase)         # Required dependency for metagenomeSeq (for AnnotatedDataFrame)
library(stringr)         # For string manipulation (e.g., trimming taxonomy strings)

#conflicts(detail = TRUE)

```

# Load the BIOM file

Load the BIOM file as an `MRexperiment` object (change depending on your biom file to be used for downstream applications)

```{r Load biom and convert to MRexperiment, include=T, echo=T}

mr_exp <- loadBiom("InputData/biom_files/050525_feature_table_w_taxonomy.biom") #Change file depending on the dataset. 

```

# Cleaning dataset prior to normalization

## Filtering Non-Prokaryote
This step keeps only features classified as Bacteria or Archaea to prevent eukaryotic/unknown reads (e.g., chloroplast/mitochondria) from biasing CSS normalization. It uses the featureData field `X1` (QIIME-style d__ prefixes) and retains rows matching `d__Bacteria` or `d__Archaea`. If your BIOM uses a different field or prefixes, update the column name/values accordingly. The chunk prints the domains that remain after filtering and updates `mr_exp` in place.

### V1 of the Non-Prokaryote filtering code

```{r Filtering non Bacteria or non Archaea, echo=T, include=T}


# Filter out features not classified as Bacteria or Archaea
# Occasionally, some Eukaryotic or unassigned sequences may be misclassified

if ("X1" %in% colnames(mr_exp@featureData@data)) {
  domain_values <- mr_exp@featureData@data$X1
  valid_rows <- domain_values %in% c("d__Archaea", "d__Bacteria")
  mr_exp <- mr_exp[valid_rows, ]
  cat("Remaining domains:", unique(mr_exp@featureData@data$X1), "\n")
}


```

### V2 of the filtering code

The following code box is a new version of the non-Bacteria taxa filtering for a more general dataset (i.e., in our case, `X1` was used. But this is not always the case). Since is a new code portion, I need to test if this works as the one above before choosing the version. 


```{r Filtering non Bacteria or non Archaea V2 (DO NOT USE YET), echo=T, include=T}

#Robust domain filtering (works whether domain is in `X1`, `kingdom`, or the `taxonomy` string)
fdat <- fData(mr_exp)

# Try common fields first; fall back to parsing a taxonomy string column if present
domain_col <- intersect(c("X1","kingdom","Domain","domain"), colnames(fdat))
if (length(domain_col) == 1) {
  dom_vals <- fdat[[domain_col]]
} else if ("taxonomy" %in% colnames(fdat)) {
  # e.g., "k__Bacteria; p__Firmicutes; ..."
  dom_vals <- sub(";.*$", "", fdat$taxonomy)         # keep first rank
} else {
  dom_vals <- NA_character_
}

keep <- rep(TRUE, nrow(fdat))
if (!all(is.na(dom_vals))) {
  keep <- grepl("(d__|k__)?Archaea|Bacteria", dom_vals, ignore.case = TRUE)
}

mr_exp <- mr_exp[keep, ]
cat("Kept features (Bacteria/Archaea filter):", sum(keep), "of", length(keep), "\n")


```


## Identify and temporarily drop zero-count samples.

CSS requires each sample to have a positive library size; columns with total counts = 0 cause failures/NaNs in normalization. Here we **(1)** capture the original sample order and phenodata, **(2)** compute per-sample sums, **(3)** list empty vs. non-empty samples, and **(4)** subset to `mr_exp_nonempty` for CSS. 

The console output reports how many samples are kept/dropped. Empty samples are only removed temporarily, as we’ll reinsert them later as zero-filled columns to preserve the original sample list/order for downstream merges.

```{r cleaning data before normalization, echo = T, include=T}

# Identify and remove empty samples (total count = 0)
# This is necessary before performing CSS normalization

orig_samples <- colnames(MRcounts(mr_exp))
orig_pheno <- pData(mr_exp)

counts <- MRcounts(mr_exp)
sample_sums <- colSums(counts)
empty_samples <- names(sample_sums)[sample_sums == 0]
non_empty_samples <- names(sample_sums)[sample_sums > 0]

cat("Total samples:", length(orig_samples), "\n")
cat("Non-empty samples:", length(non_empty_samples), "\n")
cat("Empty samples:", length(empty_samples), "\n")

mr_exp_nonempty <- mr_exp[, non_empty_samples]

```

# CSS normalization

CSS scales each sample by the cumulative sum of counts up to a chosen percentile p, reducing library-size effects in sparse data. Although `cumNormStat()` can estimate a data-driven p, here we fix p = 0.5 (median) to ensure reproducibility across runs. We normalize with `cumNorm()` and then extract the CSS-normalized counts via `MRcounts(..., norm = TRUE, log = FALSE)`. The console message confirms the percentile used. 

If you prefer automatic selection, uncomment `cumNormStat()` and pass its result to `cumNorm()`.

```{r CSS normalization, echo=T, include=T}
#NORMALIZATION
#p_est <- cumNormStat(mr_exp_nonempty) #This is the recommended, but will do manual p instead. 

manual_p <- 0.5 #normalization percentile
mr_exp_norm <- cumNorm(mr_exp_nonempty, p = manual_p)
norm_counts <- MRcounts(mr_exp_norm, norm = TRUE, log = FALSE)

cat("CSS normalization completed using manual p =", manual_p, "\n")

```

# Prepare dataset to save

## Revert old biom configuration

For stable downstream joins/BIOM round-trips, we add back the samples that were temporarily dropped for CSS because they had zero reads. 

This chunk: 

**(1)** creates zero-filled columns for `empty_samples`, 

**(2)** binds them to the normalized matrix, and 

**(3)** reorders columns to exactly match `orig_samples`. 

The resulting `full_norm_counts` thus contains CSS-normalized counts for non-empty samples and zeros for the previously empty ones. Keep in mind: analyses that require positive library sizes should still exclude all-zero samples at analysis time. A dimension printout is emitted for a quick sanity check.

```{r Revert old biom configuration, echo=T, include=T}
# Reinsert empty samples as zero-filled columns
# This step ensures consistency with the original sample order and count structure

if(length(empty_samples) > 0) {
  zeros_mat <- matrix(0, nrow = nrow(norm_counts), ncol = length(empty_samples),
                      dimnames = list(rownames(norm_counts), empty_samples))
  full_norm_counts <- cbind(norm_counts, zeros_mat)
  full_norm_counts <- full_norm_counts[, orig_samples, drop = FALSE]
} else {
  full_norm_counts <- norm_counts
}
cat("Normalized count matrix dimensions:", dim(full_norm_counts), "\n")


```

## Save normalized biom

```{r Save normalized biom, echo=T, include=T}

# Save the normalized count matrix as a new BIOM file
# This includes all samples (even those originally empty) and retains taxonomy

mr_exp_full <- newMRexperiment(full_norm_counts,
                               phenoData = AnnotatedDataFrame(orig_pheno),
                               featureData = AnnotatedDataFrame(fData(mr_exp)))

new_biom <- MRexperiment2biom(mr_exp_full, norm = FALSE)
# Note: counts in mr_exp_full are already CSS-normalized; norm=FALSE was set above so no *additional* normalization is applied during export.

#Rename the biom file as you pleased.
write_biom(new_biom, biom_file = "InputData/biom_files/050525_feature_table-css-norm_w_taxonomy.biom")

cat("CSS normalization complete. File saved with", ncol(MRcounts(mr_exp_full)), "samples.\n")
cat("The biom file is now ready for downstream use.")

```

