---
title: "Code 2: Pre_processing_After_cssNorm"
author: "Bryan Rodriguez-Colon and Yasawantha Hiripitiyage"
date: "Last udpated: 2025-09-16"
output: html_document
---

***Disclaimer: This R Markdown script was originally created by Yasawantha Hiripitiyage and has been adapted to for new data by Bryan Rodriguez Colon with some use of AI (chatGPT-4o & chatGPT-5)***

This R Markdown script is used to turn the biom, metadata and phylogenetic tree into `phyloseq` objects for downstream microbial ecology analysis.

**It performs the following tasks:**

1. Loads CSS-normalized and non-normalized BIOM files.
2. Parses taxonomy formatted using **SILVA 138** conventions.
3. Loads and formats associated metadata and phylogenetic tree.
4. Combines abundance, taxonomy, metadata, and tree into a single `phyloseq` object.
5. It created a phyloseq object with the non-nomralized biom for code 6 (statistical analysis).
6. Subsets the data by domain: **Bacteria** and **Archaea**.
7. Saves the resulting `phyloseq` objects as `.RData` files for future analysis.



```{r setup, include=T}
# Prepare script
knitr::opts_chunk$set(echo = TRUE)

setwd("/Users/bryanrodriguez/Library/CloudStorage/OneDrive-UniversityofKansas/Bryan-MS_PhD/PHD-PROJECT/PuertoRico/CHAPTER2/16S/Processing/Geobio_Guanica_SciRep_Codes/R_Codes")

rm(list = ls())


```


# Load libraries
```{r load packages, message=F, include=T}
# Load required package for microbiome data integration
library(phyloseq)       # For microbiome data manipulation and visualization
library(dplyr)          # For data manipulation (e.g., group_by, summarize)
library(biomformat)     # For reading/writing BIOM files
```

# Functions

## parse_taxonomy_silva138
Parses taxonomic strings formatted with SILVA 138 prefixes (e.g., `d__Bacteria`; `p__Proteobacteria`; …). It maps one-letter prefixes to standard ranks (Domain, Phylum, Class, Order, Family, Genus, Species), removes the prefixes, and guarantees a consistent 7-level output by filling missing ranks with `NA`. This standardization is required for clean import into `phyloseq` (e.g., when building a tax_table) from QIIME2/SILVA outputs.

**Steps:**

• Removes SILVA-style prefixes (e.g., `d__`, `p__`, etc.; `k__` accepted as Domain).

• Assigning standard rank names to each level.

• Filling in missing taxonomic ranks with `NA` to ensure a complete 7-level taxonomy.

This standardized output is required for correct parsing when importing taxonomy into phyloseq using QIIME2 outputs and SILVA 138 formatting.

@param `char.vec` A single semicolon-delimited taxonomy string (e.g., from a BIOM file).

@return A named character vector of length 7 (`Domain`, `Phylum`, `Class`, `Order`, `Family`, `Genus`, `Species`) with cleaned names and `NA` where missing.

*Note: Uses `parse_taxonomy_default()` to split the string; issues a warning if no SILVA-style prefixes are detected.*

```{r function parse_taxonomy_silva138, include=T, echo=T}
# Function to parse SILVA 138-style taxonomies into standard ranks
# Converts d__, p__, etc. to Domain, Phylum, etc.
parse_taxonomy_silva138 <- function(char.vec) {
  char.vec <- parse_taxonomy_default(char.vec)
  Tranks <- c(d = "Domain", p = "Phylum", c = "Class", 
              o = "Order", f = "Family", g = "Genus", 
              s = "Species")
  ti <- grep("[[:alpha:]]{1}\\_\\_", char.vec)
  if (length(ti) == 0L) {
    warning("No SILVA138 prefixes were found. \n", 
            "Consider using parse_taxonomy_default() instead if true for all ASVs/OTUs. \n")
    taxvec <- char.vec
  } else {
    taxvec <- gsub("[[:alpha:]]{1}\\_\\_", "", char.vec)
    repranks <- Tranks[substr(char.vec[ti], 1, 1)]
    names(taxvec)[ti[!is.na(repranks)]] <- repranks[!is.na(repranks)]
  }
  desired_ranks <- c("Domain", "Phylum", "Class", "Order", "Family", "Genus", "Species")
  taxvec <- taxvec[names(taxvec) %in% desired_ranks]
  if(length(taxvec) != length(desired_ranks)) {
    new_tax <- setNames(rep(NA, length(desired_ranks)), desired_ranks)
    new_tax[names(taxvec)] <- taxvec
    taxvec <- new_tax
  }
  return(taxvec)
}


```


# Creating the Phyloseq Objects

## Load BIOMs into phyloseq (raw vs. CSS-normalized).

Here we import two versions of the same table: (1) the **non-normalized** BIOM (kept as integer counts; to use this for statistical steps such as Alpha diversity), and (2) the **CSS-normalized** BIOM produced in Code1. Both imports use `parse_taxonomy_silva138` so ranks are standardized (`Domain…Species`). The two phyloseq objects share identical ASV IDs, sample IDs, and taxonomy (only the abundance values differ). 

Be explicit downstream about which object you use for each analysis step. For example, to compute beta diversity, use CSS-normalized data instead of non-normalized, (i.e., `Arch_Bac_norm_biom`).


```{r loading abundance data and phylo data to phyloseq, include = T, echo = T}

# Import the non-normalized biom
# The non normalized biom will be used for statistical analysis (e.g., alpha diversity)
Arch_Bac_biom <- import_biom("InputData/biom_files/050525_feature_table_w_taxonomy.biom",
                                         parseFunction = parse_taxonomy_silva138)

# Import the CSS-normalized BIOM table and parse taxonomy using the SILVA 138 function
Arch_Bac_norm_biom <- import_biom("InputData/biom_files/050525_feature_table-css-norm_w_taxonomy.biom",
                                         parseFunction = parse_taxonomy_silva138)

```


### Checking Rank names of data
```{r check rank names}

# Optional: View the taxonomy ranks available in the dataset
rank_names(Arch_Bac_norm_biom)

```


## Load sample metadata and convert to `phyloseq::sample_data`.

We read the project metadata from CSV, then write a QIIME-style, tab-delimited mapping file because `import_qiime_sample_data()` expects that format. `check.names = FALSE` preserves column names exactly (helpful if you use spaces or symbols), and `stringsAsFactors = FALSE` prevents unintended factor coercion. The importer returns a `sample_data` object (named `Arch_Bac_map`) ready to merge with your phyloseq objects.

*Note: The mapping file should have the first column as the sample identifier (ideally #SampleID in QIIME format) and those IDs must match the BIOM sample names.*

```{r load map file data to R}

# Load the mapping file (metadata) and convert it to `sample_data` object
meta <- read.csv("InputData/SciRep_metadata_filtered.csv", header = TRUE, stringsAsFactors = FALSE, check.names = FALSE)
write.table(meta, file = "InputData/SciRep_metadata_filtered.txt", sep = "\t", row.names = FALSE, quote = FALSE)
Arch_Bac_map <- import_qiime_sample_data("InputData/SciRep_metadata_filtered.txt")
rm(meta) #removing variables to clean environment


```

## Loading Phylogenetic Tree to R

Loads the 16S phylogenetic tree in Newick format using `phyloseq::read_tree()`. Tip labels (ASV/OTU IDs) must **exactly match** the feature IDs in your BIOM objects to enable UniFrac, Faith’s PD, and other phylogeny-aware methods. If the tree is unrooted, consider midpoint or outgroup rooting before use.

```{r load tree file data to R}

# Load tree file in Newick format
Arch_Bac_tree <- read_tree("InputData/050525_tree.nwk")

```


## Assemble the normalized phyloseq object (abundance + taxonomy + metadata + tree).
We merge the CSS-normalized BIOM (`Arch_Bac_norm_biom`), the sample metadata (`Arch_Bac_map`), and the phylogenetic tree (`Arch_Bac_tree`) into a single object (`Arch_Bac_norm`). `merge_phyloseq()` intersects on both sample names and taxa IDs—any non-matching entries are silently dropped—so ensure IDs are synchronized beforehand. 

```{r combine data, echo=F, include=F}

# Merge BIOM table, metadata, and tree into one `phyloseq` object
Arch_Bac_norm <- merge_phyloseq(Arch_Bac_norm_biom,
                                       Arch_Bac_map,
                                       Arch_Bac_tree)
Arch_Bac_norm

```


### We will now seperate Normalized Archaea & Bacteria into two seperate datasets. 

This is in case you want to only plot Bacteria or Archaea. 

```{r subset archaea and bacteria}
#Subset archaea
Arch_norm <- subset_taxa(Arch_Bac_norm, Domain == "Archaea")
Arch_norm

#Subset bacteria
Bac_norm <- subset_taxa(Arch_Bac_norm, Domain == "Bacteria")
Bac_norm
```


# Non-Normalized phyloseqs (for Alpha Diversity)

This Phyloseq with the non-normalized data is constructed for alpha diversity analysis (code 6). 

```{r}
# Merge BIOM table, metadata, and tree into one `phyloseq` object
Arch_Bac <- merge_phyloseq(Arch_Bac_biom,
                                       Arch_Bac_map,
                                       Arch_Bac_tree)
Arch_Bac

#Subset archaea
Arch <- subset_taxa(Arch_Bac, Domain == "Archaea")
Arch

#Subset bacteria
Bac <- subset_taxa(Arch_Bac, Domain == "Bacteria")
Bac


```




# Saving RDatafiles

We are going to save data as RDatafiles so that we can gain easy access.


```{r save as RData}
save(Arch_Bac_norm, file = "OutputData/RData/SciRep_Arch_Bac_norm_050525.RData")
save(Arch_norm, file = "OutputData/RData/SciRep_Arch_norm_050525.RData")
save(Bac_norm, file = "OutputData/RData/SciRep_Bac_norm_050525.RData")

cat("Data is ready for code 3 (relative abundance calculations)")

save(Arch_Bac, file = "OutputData/RData/SciRep_Arch_Bac_050525.RData")
save(Arch, file = "OutputData/RData/SciRep_Arch_050525.RData")
save(Bac, file = "OutputData/RData/SciRep_Bac_050525.RData")

cat("Data is ready for Alpha Diversity in code 6 (statistial analysis)")

```






