---
title: 'Code 7: taxonomy Summaries'
author: "Bryan Rodriguez"
date: "2025-05-16"
output: html_document
---

This script uses the prepared melted relative abundance tables, and outputs **Excel files summarizing taxonomy** at different levels (Phylum to Genus) for:

- All Prokaryotes  
- Bacteria only  
- Archaea only  

Summaries are generated across three grouping variables:

- `Sample_Layer`
- `Sample_name`
- `Sample_Type`



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

setwd("/Users/bryanrodriguez/Library/CloudStorage/OneDrive-UniversityofKansas/Bryan-MS_PhD/PHD-PROJECT/PuertoRico/CHAPTER2/16S/Processing/Geobio_Guanica_Codes/R_Codes")

rm(list = ls())


```



# loading libraries
```{r load packages, echo=F, include=F}
# Load required packages
library(phyloseq)     # For microbiome data structures and manipulation
library(data.table)   # For efficient data handling
library(dplyr)        # For data manipulation (e.g., group_by, summarize)
library(openxlsx)     # To write Excel files

```



# Functions

## Function: save_complete_taxa_summaries()

Generates and saves detailed Excel summaries of taxonomic composition from a melted relative abundance data frame. It outputs:

One summary file including all taxonomic levels (Phylum to Genus) grouped by a selected metadata variable.
Separate summary files for each individual taxonomic rank. Files are saved in organized folders by microbial group (e.g., Bacteria, Archaea) and grouping variable (e.g., Sample_name, Sample_Layer).


```{r Function to save summary into excel files,echo=F}

save_complete_taxa_summaries <- function(melted_df, taxa_ranks, group_name, group_var, output_dir = "DataSummary") {

  # Ensure the main output directory exists
  if (!dir.exists(output_dir)) dir.create(output_dir, recursive = TRUE)

  # Create subdirectory based on group name
  subfolder_path <- file.path(output_dir, group_name)
  if (!dir.exists(subfolder_path)) dir.create(subfolder_path, recursive = TRUE)

  # Create a subfolder for the grouping variable
  grouping_folder <- file.path(subfolder_path, group_var)
  if (!dir.exists(grouping_folder)) dir.create(grouping_folder, recursive = TRUE)

  # Define the specific ordering for each group_var
  sample_order <- switch(group_var,
                         "Sample_Layer" = c("POE1-gc", "POE1-g", "POE1-R", "POE1-B", 
                                            "JB-C", "JB-g", "JB-DP", "JB-B", 
                                            "PS-C1", "PS-C2", "PS-Br", "PS-L", "PS-Bottom"),
                         "Sample_name" = c("POE1", "JB", "PS"),
                         "Sample_Type" = c("Microbialite", "Microbial_Mat"),
                         unique(melted_df[[group_var]])  # fallback if no match
  )

  # Overall Summary Excel
  cat("Creating Overall Summary Excel for", group_name, "grouped by", group_var, "...\n")

  overall_summary <- melted_df %>%
    dplyr::group_by(.data[[group_var]], Phylum, Class, Order, Family, Genus) %>%
    dplyr::summarize(Rel_Abund = sum(RelAbund), Rel_Abund_SD = sd(RelAbund), .groups = "drop") %>%
    dplyr::arrange(.data[[group_var]], desc(Rel_Abund)) %>%
    dplyr::filter(Rel_Abund != 0)

  overall_list <- split(overall_summary, overall_summary[[group_var]])

  # Sort the sheets according to the predefined order
  overall_list <- overall_list[sample_order[sample_order %in% names(overall_list)]]

  overall_wb <- createWorkbook()

  for (p in names(overall_list)) {
    addWorksheet(overall_wb, sheetName = p)
    writeData(overall_wb, sheet = p, overall_list[[p]])
  }

  overall_file <- file.path(grouping_folder, paste0(group_name, "_Summary_by_", group_var, ".xlsx"))
  saveWorkbook(overall_wb, file = overall_file, overwrite = TRUE)
  cat("Saved Overall Summary:", overall_file, "\n")

  # Individual Excel Files per Taxonomic Rank
  cat("Creating Per-Taxonomic Rank Excel Files for", group_name, "grouped by", group_var, "...\n")

  for (taxa_rank in taxa_ranks) {

    summarized_data <- melted_df %>%
      dplyr::group_by(.data[[group_var]], .data[[taxa_rank]]) %>%
      dplyr::summarize(Rel_Abund = sum(RelAbund), Rel_Abund_SD = sd(RelAbund), .groups = "drop") %>%
      dplyr::arrange(.data[[group_var]], desc(Rel_Abund)) %>%
      dplyr::filter(Rel_Abund != 0)

    taxa_list <- split(summarized_data, summarized_data[[group_var]])

    # Sort the sheets according to the predefined order
    taxa_list <- taxa_list[sample_order[sample_order %in% names(taxa_list)]]

    wb <- createWorkbook()

    for (p in names(taxa_list)) {
      addWorksheet(wb, sheetName = p)
      writeData(wb, sheet = p, taxa_list[[p]])
    }

    file_name <- file.path(grouping_folder, paste0(group_name, "_Summary_", taxa_rank, "_by_", group_var, ".xlsx"))
    saveWorkbook(wb, file = file_name, overwrite = TRUE)

    cat("Saved:", file_name, "\n")
  }
}



```


```{r To reset data without erasing functions}
#To reset data without erasing functions
rm(list = ls()[!sapply(ls(), function(x) is.function(get(x)))])
```



# Load Data

```{r}
#Load data by sample layer


SLayer_LIST <- readRDS("OutputData/RData/PE_GuanicaGeobio_Arch_Bac_merged_rel_melt_SLayer.RDS")

Arch_Bac_merged_rel_melt_SLayer <- SLayer_LIST$Arch_Bac_merged_rel_melt_SLayer
Arch_merged_rel_melt_SLayer <- SLayer_LIST$Arch_merged_rel_melt_SLayer
Bac_merged_rel_melt_SLayer <- SLayer_LIST$Bac_merged_rel_melt_SLayer


#Load Data by Sample name
SName_LIST <- readRDS("OutputData/RData/PE_GuanicaGeobio_Arch_Bac_merged_rel_melt_SName.RDS")

Arch_Bac_merged_rel_melt_SName <- SName_LIST$Arch_Bac_merged_rel_melt_SName
Arch_merged_rel_melt_SName <- SName_LIST$Arch_merged_rel_melt_SName
Bac_merged_rel_melt_SName <- SName_LIST$Bac_merged_rel_melt_SName

#Load data by sample type
SType_LIST <- readRDS("OutputData/RData/PE_GuanicaGeobio_Arch_Bac_merged_rel_melt_SType.RDS")

Arch_Bac_merged_rel_melt_SType <- SType_LIST$Arch_Bac_merged_rel_melt_SType
Arch_merged_rel_melt_SType <- SType_LIST$Arch_merged_rel_melt_SType
Bac_merged_rel_melt_SType <- SType_LIST$Bac_merged_rel_melt_SType

rm(SLayer_LIST, SName_LIST, SType_LIST)

```



# Taxonomy Summary for Manuscript Results. 

## Define Taxonomic Ranks
```{r taxonomic Ranks}
# Define taxonomic ranks to process
taxa_ranks <- c("Domain","Phylum", "Class", "Order", "Family", "Genus")
```


## Process samples based on Sample_Layer
```{r Summary based on Sample Layer}

#Run Summary Function and save as excel file
save_complete_taxa_summaries(Arch_Bac_merged_rel_melt_SLayer, taxa_ranks, group_name = "All_Prokaryotes", group_var = "Sample_Layer")

#Run Summary Function and save as excel file
save_complete_taxa_summaries(Arch_merged_rel_melt_SLayer, taxa_ranks, group_name = "Archaea", group_var = "Sample_Layer")

#Run Summary Function and save as excel file
save_complete_taxa_summaries(Bac_merged_rel_melt_SLayer, taxa_ranks, group_name = "Bacteria", group_var = "Sample_Layer")


```


## Process Samples based on Sample_name
```{r Summary based on Sample name}

#Run Summary Function and save as excel file
save_complete_taxa_summaries(Arch_Bac_merged_rel_melt_SName, taxa_ranks, group_name = "All_Prokaryotes", group_var = "Sample_name")


#Run Summary Function and save as excel file
save_complete_taxa_summaries(Arch_merged_rel_melt_SName, taxa_ranks, group_name = "Archaea", group_var = "Sample_name")


#Run Summary Function and save as excel file
save_complete_taxa_summaries(Bac_merged_rel_melt_SName, taxa_ranks, group_name = "Bacteria", group_var = "Sample_name")


```


## Process Samples based on sample type
```{r Summary based on Sample type}

#Run Summary Function and all save as excel file
save_complete_taxa_summaries(Arch_Bac_merged_rel_melt_SType, taxa_ranks, group_name = "All_Prokaryotes", group_var = "Sample_Type")

#Run Summary Function on Archaea and save as excel file
save_complete_taxa_summaries(Arch_merged_rel_melt_SType, taxa_ranks, group_name = "Archaea", group_var = "Sample_Type")

#Run Summary Function on Bacteria and save as excel file
save_complete_taxa_summaries(Bac_merged_rel_melt_SType, taxa_ranks, group_name = "Bacteria", group_var = "Sample_Type")


print("All taxonomic summaries are stored and ready for the results!")
```














