---
title: "Code 5: MicrobialGuilds_plotting"
author: "Bryan Rodriguez-Colon and Yasawantha Hiripitiyage"
date: "Last udpated: 2025-09-17"
output: html_document
---

***Disclaimer: This R Markdown script was originally created by Yasawantha Hiripitiyage and has been adapted to for new data by Bryan Rodriguez Colon with some use of AI (chatGPT-4o & chatGPT-5).***

## Overview

This R Markdown generates manuscript figures summarizing the relative abundance of selected **microbial functional guilds** across samples, primarily grouped by `Sample_Layer`. The workflow uses previously prepared long-format relative abundance tables and aggregates them by taxonomic rank before subsetting taxa into curated functional guilds (e.g., Cyanobacteria, sulfate reducers, methanogens, purple sulfur bacteria, purple non-sulfur bacteria, sulfur oxidizers, and additional optional metabolic groups).

Guild-specific datasets are constructed using taxon-level filters and summarized as relative abundance percentages. The script produces stacked bar plots (family-level fill in the current implementation) with fixed sample ordering and reproducible color palettes to ensure consistency across figures. These visualizations are intended to highlight functional differences among microbial communities across depositional settings (Supplementary Figure S7).




```{r Setup, include=T}
knitr::opts_chunk$set(echo = TRUE)
rm(list = ls())
setwd("/Users/bryanrodriguez/Library/CloudStorage/OneDrive-UniversityofKansas/Bryan-MS_PhD/PHD-PROJECT/PuertoRico/CHAPTER2/16S/Processing/Geobio_Guanica_SciRep_Codes/R_Codes")

```


# Loading libraries
```{r Loading packages, message=F, include=T}
library(dplyr)        # used for group_by/summarize/arrange/filter/mutate/bind_rows/ungroup %>% pipes
library(tibble)       # used via add_column()
library(ggplot2)      # all plotting uses ggplot2
library(RColorBrewer) # brewer.pal() used to build the combined qualitative palette
library(cowplot)      # cowplot::get_legend() used to extract legends
library(tidyr)        # complete() used to fill missing Sample_Layer × Family combos with zeros

```


## Stop data displayed in scientific notation

```{r stop scipen}
options(scipen = 999)
```



# Functions

## Helper Function to Extract Legend: 'extract_legend()'



```{r Helper Function to Extract Legend}

# Function to extract just the legend from a ggplot
extract_legend <- function(plot) {
  g <- ggplotGrob(plot)
  legend_index <- which(sapply(g$grobs, function(x) x$name) == "guide-box")
  legend <- g$grobs[[legend_index]]
  return(legend)
}

```



# Data
Reading previously saved melt Data
```{r load saved melt data}
melt_file_list <- readRDS("OutputData/RData/PE_GuanicaGeobio_Arch_Bac_merged_rel_melt_SLayer.RDS")



Arch_Bac_merged_rel_melt <- melt_file_list$Arch_Bac_merged_rel_melt_SLayer

Arch_merged_rel_melt <- melt_file_list$Arch_merged_rel_melt_SLayer
Bac_merged_rel_melt <- melt_file_list$Bac_merged_rel_melt_SLayer
rm(melt_file_list)
```



# Preparing data

## Splitting Data
Splitting Data by Sample Layer summarizing abundance data at each taxonomic level and sorting abundance in descending order. 
```{r split data by sample layer}
Arch_merged_rel_melt_group <- Arch_merged_rel_melt %>% dplyr::group_by(Sample_Layer, Phylum, Class, Order, Family, Genus, Species) %>% dplyr::summarize(Rel_Abund = sum(RelAbund), Rel_Abund_SD = sd(RelAbund)) %>% arrange(desc(Rel_Abund))

Bac_merged_rel_melt_group <- Bac_merged_rel_melt %>% dplyr::group_by(Sample_Layer, Phylum, Class, Order, Family, Genus, Species) %>% dplyr::summarize(Rel_Abund = sum(RelAbund), Rel_Abund_SD = sd(RelAbund)) %>% arrange(desc(Rel_Abund))

Arch_Bac_merged_rel_melt_group <- Arch_Bac_merged_rel_melt %>% dplyr::group_by(Sample_Layer, Phylum, Class, Order, Family, Genus, Species) %>% dplyr::summarize(Rel_Abund = sum(RelAbund), Rel_Abund_SD = sd(RelAbund)) %>% arrange(desc(Rel_Abund))

rm(Arch_merged_rel_melt, Bac_merged_rel_melt, Arch_Bac_merged_rel_melt)

```

# Subset Data by microbial guilds

## Main Guilds to check

### Subset Cyanobacteria

```{r subset cyanobacteria}
Cyano <- Arch_Bac_merged_rel_melt_group %>%
  filter(
    Phylum == "Cyanobacteriota" |
    Family %in% c("Cyanobacteriia_Unclassified", "Oscillatoriaceae", "Coleofasciculaceae", "Leptolyngbyaceae", "Cyanobiaceae", "Nodosilineaceae", "Desertifilaceae"),
    Rel_Abund != 0,
    Rel_Abund >= 0.5
  ) %>%
  add_column(Guild = "Cyanobacteria")

Cyano
```

### Subset Purple non-sulfur bacteria (PNSB)
```{r subset purple non sulf bac}


pnsb_genera <- c(
  "Rhodovulum",
  "Rhodovibrio",
  "Rhodothalassium",
  "Rhodobacter",
  "Rhodopseudomonas",
  "Rhodospirillum",
  "Rhodomicrobium",
  "Rhodoplanes",
  "Rhodocista",
  "Rhodobium",
  "Blastochloris",
  "Rhodobaca",
  "Roseospira",
  "Roseospirillum",
  "Phaeospirillum",
  "Rhodopila",
  "Rhodospira",
  "Rhodoblastus",
  "Caenispirillum"
)

# Filter ONLY these genera, not whole families or orders
PNSB <- Arch_Bac_merged_rel_melt_group %>%
  filter(
    Genus %in% pnsb_genera,
    Rel_Abund != 0,
    Rel_Abund >= 0.5
  ) %>%
  mutate(Guild = "Purple Non-Sulfur Bacteria")
PNSB

```



### Sulfate Reducing Bacteria (SRB)

```{r subset SRB}

srb_families <- c(
  "Desulfovibrionaceae", "Desulfobacteraceae", "Desulfohalobiaceae",
  "Desulfobulbaceae", "Desulfotomaculaceae", "Thermodesulfobacteriaceae",
  "Archaeoglobaceae", "Desulfitibacteraceae", "Desulfomonilaceae",
  "Desulfatibacillaceae", "Desulfarculaceae", "Desulfococcaceae",
  "Syntrophobacteraceae", "Desulfobaccaceae", "Desulfatirhabdiaceae", "Desulfatiglandaceae", "Desulfosarcinaceae", "Desulfomonilales", "Desulfobacterales"
)



SRB <- Arch_Bac_merged_rel_melt_group %>%
  filter(Family %in% srb_families & Rel_Abund != 0, Rel_Abund >= 0.5) %>%
  add_column(Guild = "Sulfate Reducing")
SRB


```

### Subset Methanogens

```{r subset methanogens}

methanogen_families <- c(
  "Methanosarcinaceae", "Methanomassiliicoccaceae", "Methanobacteriaceae",
  "Methanococcaceae", "Methanocaldococcaceae", "Methanomicrobiaceae",
  "Methanosaetaceae", "Methanoregulaceae", "Methanotrichaceae",
  "Methanofastidiosales_Unclassified", "Methanonatronarchaeaceae",
  "Methanobacteriota_Unclassified"
)

MET <- Arch_Bac_merged_rel_melt_group %>%
  filter(Family %in% methanogen_families & Rel_Abund != 0, Rel_Abund >= 0.5) %>%
  add_column(Guild = "Methanogens")
MET

```


## Other Optional Guilds

### Subset Green non-sulfur Bacteria (GNSB)

```{r subset green non-sulfur bac}
#### Green Non-Sulfur Bacteria (GNSB)

gnsb_genera <- c("Chloroflexus", "Roseiflexus", "Oscillochloris", "Herpetosiphon")

GNSB <- Arch_Bac_merged_rel_melt_group %>%
  filter(
    Genus %in% gnsb_genera |
    Order %in% c("Chloroflexales", "Herpetosiphonales") |
    Family %in% c("Chloroflexaceae", "Anaerolineaceae", "Thermoflexaceae", "Caldilineaceae", "Acetothermiia_Unclassified", "Chloroflexota_Unclassified"),
    Rel_Abund != 0,
    Rel_Abund >= 0.5
  ) %>%
  mutate(Guild = "Green Non-Sulfur Bacteria")
GNSB
```


### Subset Purple Sulfur bacteria (PSB)
```{r subset purple sulfer bac}
#### Purple Sulfur Bacteria (PSB)

psb_genera <- c(
  "Allochromatium", "Amoebobacter", "Chromatium", "Halochromatium", 
  "Isochromatium", "Lamprocystis", "Lamprobacter", "Marichromatium", 
  "Rhabdochromatium", "Thermochromatium", "Thioalkalicoccus", "Thiobaca", 
  "Thiocapsa", "Thiococcus", "Thiocystis", "Thiodictyon", "Thioflavicoccus", 
  "Thiohalocapsa", "Thiolamprovum", "Thiopedia", "Thiorhodococcus", 
  "Thiorhodovibrio", "Thiospirillum", "Halorhodospira", "Thiorhodospira", 
  "Ectothiorhodosinus"
)

PSB <- Arch_Bac_merged_rel_melt_group %>%
  filter(
    Genus %in% psb_genera,
    Rel_Abund != 0,
    Rel_Abund >= 0.5
  ) %>%
  mutate(Guild = "Purple Sulfur Bacteria")
PSB

```


### Subset Sulfur Oxidizers (SOx)

```{r subset sulfur oxidizers}


sox_genera <- c(
  "Thiomicrospira", "Beggiatoa", "Thiothrix", "Thiomargarita", 
  "Thioalkalivibrio", "Thiomonas", "Paracoccus", "Sulfitobacter", 
  "Sulfurovum", "Sulfurimonas", "Thiobacillus", "Acidithiobacillus", 
  "Halothiobacillus", "Thioalkalicoccus", "Thiohalospira",
  "Guyparkeria", "Thiomicrohabdus", "Thiohalophilus",
  "Candidatus Allobeggiatoa"
)

SOx <- Bac_merged_rel_melt_group %>%
  filter(
    Genus %in% sox_genera |
    Genus %in% psb_genera,
    Rel_Abund != 0,
    Rel_Abund >= 0.5
  ) %>%
  mutate(Guild = "Sulfur Oxidizers")
SOx

```


### Subset Iron Reducers (FeR)
```{r subset iron reducers}
#### Iron Reducers

iron_reducer_families <- c("Geobacteraceae", "Desulfuromonadales_Unclassified")

FeR <- Bac_merged_rel_melt_group %>%
  filter(Family %in% iron_reducer_families & Rel_Abund != 0, Rel_Abund >= 0.5) %>%
  mutate(Guild = "Iron Reducers")
FeR
```

### Subset Ammonia Oxidizers (NH3Ox)


```{r subset Ammonia Oxidizers}

nitrogen_AmmoniaOx_families <- c("Nitrosococcaceae", "Nitrososphaeraceae", "Nitrococcaceae")

NH3Ox <- Arch_Bac_merged_rel_melt_group %>%
  filter(Family %in% nitrogen_AmmoniaOx_families & Rel_Abund != 0, Rel_Abund >= 0.5) %>%
  mutate(Guild = "Ammonia Oxidizers")
NH3Ox
```

### Subset Denitrifiers (DEN)

```{r subset denitrifiers}
#### Denitrifiers

denitrifier_genera <- c("Achromobacter", "Aerobacter", "Alcaligenes", "Arcobacter", "Aromatoleum", "Azoarcus", "Bacillus", "Castellaniella", "Comamonas", "Curvibacter", "Dechloromonas", "Enterobacter", "Flavobacterium", "Georgfuchsia", "Hyphomicrobium", "Meganema", "Methylobacillus", "Methylophilus", "Micrococcus", "Paracoccus", "Proteus", "Pseudomonas", "Rhodobacter", "Rhodococcus", "Serratia", "Simplicispira", "Sulfuritalea", "Thauera", "Thiobacillus")

DEN <- Bac_merged_rel_melt_group %>%
  filter(Genus %in% denitrifier_genera & Rel_Abund != 0, Rel_Abund >= 0.5) %>%
  mutate(Guild = "Denitrifiers") 
DEN
```


### Subset Nitrogen Fixers (NFix)

```{r subset nitrogen fixers}
#### Nitrogen Fixers

nitrogen_fixing_genera <- c("Azospirillum", "Azotobacter", "Beijerinckia", "Bradyrhizobium", "Clostridium", "Frankia", "Klebsiella", "Mesorhizobium", "Nostoc", "Rhizobium", "Anabaena", "Desulfovibrio", "Calothrix")

NFix <- Bac_merged_rel_melt_group %>%
  filter(Genus %in% nitrogen_fixing_genera & Rel_Abund != 0, Rel_Abund >= 0.5) %>%
  mutate(Guild = "Nitrogen Fixers")
NFix
```


### Subset Aerobic Heterotrophs (AHET)


```{r subset aerobic heterotrophs}
#### Strictly Aerobic Heterotrophs

aerobic_heterotrophs_genera <- c("Pseudomonas", "Flavobacterium", "Alteromonas", "Vibrio", "Shewanella", "Myxococcus", "Sphingomonas", "Roseobacter")

AHET <- Bac_merged_rel_melt_group %>%
  filter(Genus %in% aerobic_heterotrophs_genera & Rel_Abund != 0, Rel_Abund >= 0.5) %>%
  mutate(Guild = "Strictly Aerobic Heterotrophs")
AHET
```

### Subset Methanotrophs
```{r subset methanotrophs}

#### Methanotrophs

methanotroph_genera <- c("Methylomonas", "Methylococcus", "Methylobacter", "Methylocystis", "Methylocapsa")

Methanotrophs <- Bac_merged_rel_melt_group %>%
  filter(Genus %in% methanotroph_genera & Rel_Abund != 0) %>%
  mutate(Guild = "Methanotrophs")
Methanotrophs
```





# Plotting

## Prepare for plotting
Setting ggplot2 theme
```{r set ggplot2 theme}
theme_set(theme_bw(base_size = 14))
```

Combining All Guilds
```{r combine alll guilds}
All_Guilds <- bind_rows(list(MET, SRB, Cyano, PSB, PNSB, GNSB, SOx, NH3Ox, DEN, AHET))

All_Guilds

```


Removing groups
```{r remove groups}

All_Guilds_ungroup <- ungroup(All_Guilds)
```

## Plotting by sample layer


```{r}
# Get unique genera or family, sorted for consistency
unique_GF <- sort(unique(All_Guilds_ungroup$Family))

# Combine multiple qualitative palettes
base_colors <- c(
  brewer.pal(9, "Set1"),
  brewer.pal(8, "Set2"),
  brewer.pal(8, "Dark2"),
  brewer.pal(12, "Paired")
)

# Expand as needed
color_GF_palette <- colorRampPalette(base_colors)(length(unique_GF))

# Name each color by genus (important!)
names(color_GF_palette) <- unique_GF



```

The code below is to plot the figure
```{r good plot}


# Sample layer order
sample_order_All <- c("PS-Bottom", "PS-L", "PS-Br", "PS-C2", "PS-C1", 
                      "JB-B", "JB-DP", "JB-g", "JB-C", 
                      "POE1-B", "POE1-R", "POE1-g", "POE1-gc")

# Factor Sample_Layer
All_Guilds_ungroup$Sample_Layer <- factor(All_Guilds_ungroup$Sample_Layer, levels = sample_order_All)

# Filter out families with <0.4% mean relative abundance
All_Guilds_ungroup <- All_Guilds_ungroup %>%
  group_by(Family) %>%
  filter(mean(Rel_Abund, na.rm = TRUE) >= 0) %>%
  ungroup()

# Set Family factor levels using the color palette
All_Guilds_ungroup$Family <- factor(All_Guilds_ungroup$Family, levels = names(color_GF_palette))

# Loop over guilds
guilds <- unique(All_Guilds_ungroup$Guild)

for (g in guilds) {
  guild_df <- All_Guilds_ungroup %>%
    filter(Guild == g) %>%
    complete(Sample_Layer = factor(sample_order_All, levels = sample_order_All),
             Family = factor(levels(All_Guilds_ungroup$Family)),
             fill = list(Rel_Abund = 0))

  # Build plot with legend
  p_full <- ggplot(guild_df, aes(x = Sample_Layer, y = Rel_Abund, fill = Family)) +
    geom_bar(stat = "identity") +
    coord_flip() +
    scale_fill_manual(values = color_GF_palette, drop = TRUE) +
    labs(
      title = g,
      x = "Sample Layer",
      y = "Relative abundance (% of ASVs per sample)"
    ) +
    theme_bw(base_size = 14) +
    theme(
      plot.title = element_text(hjust = 0.5, face = "bold"),
      axis.text.x = element_text(size = 10),
      axis.text.y = element_text(size = 10),
      legend.position = "right"
    )

 # Extract guild-specific legend
  legend_grob <- cowplot::get_legend(p_full)

  # Plot without legend
  p_nolegend <- p_full + theme(legend.position = "none")

  # Clean guild name for file-safe use
  file_safe_g <- gsub(" ", "_", g)

  # Save main plot
  ggsave(
    filename = paste0("plots/Microbial_Guilds/Family_Level/Fam_guild_", file_safe_g, ".svg"),
    plot = p_nolegend,
    width = 11.78,
    height = 10,
    units = "in",
    dpi = 300
  )

  ggsave(
    filename = paste0("plots/Microbial_Guilds/Family_Level/Fam_guild_", file_safe_g, ".tiff"),
    plot = p_nolegend,
    width = 11.77,
    height = 10,
    units = "in",
    dpi = 600,
    device = "tiff"
  )
  
}
```

The code below is to plot the legends 
```{r good legend}


# Define threshold
abundance_threshold <- 0

# Ensure sample order is set
sample_order_All <- c("PS-Bottom", "PS-L", "PS-Br", "PS-C2", "PS-C1", 
                      "JB-B", "JB-DP", "JB-g", "JB-C", 
                      "POE1-B", "POE1-R", "POE1-g", "POE1-gc")
All_Guilds_ungroup$Sample_Layer <- factor(All_Guilds_ungroup$Sample_Layer, levels = sample_order_All)

# Set Family factor levels to palette order
All_Guilds_ungroup$Family <- factor(All_Guilds_ungroup$Family, levels = names(color_GF_palette))

# Get unique guilds
guilds <- unique(All_Guilds_ungroup$Guild)

# Loop through guilds
for (g in guilds) {
  guild_df <- All_Guilds_ungroup %>%
    filter(Guild == g, Rel_Abund >= abundance_threshold)

  # Set Family factor again to only what's present
  guild_df$Family <- factor(guild_df$Family, levels = names(color_GF_palette))

  # Create full plot with legend
  p_full <- ggplot(guild_df, aes(x = Sample_Layer, y = Rel_Abund, fill = Family)) +
    geom_bar(stat = "identity") +
    coord_flip() +
    scale_fill_manual(values = color_GF_palette, drop = TRUE) +
    labs(
      title = g,
      x = "Sample Layer",
      y = "Relative abundance (% of ASVs per sample)"
    ) +
    theme_bw(base_size = 14) +
    theme(
      plot.title = element_text(hjust = 0.5, face = "bold"),
      axis.text.x = element_text(size = 10),
      axis.text.y = element_text(size = 10),
      legend.position = "right"
    )

  # Extract guild-specific legend
  legend_grob <- cowplot::get_legend(p_full)

  # Remove legend from main plot
  p_nolegend <- p_full + theme(legend.position = "none")

  # Save plot
  file_safe_g <- gsub(" ", "_", g)
  #ggsave(paste0("plots/Microbial_Guilds/Fam_guild_", file_safe_g, ".svg"),
         #plot = p_nolegend, width = 15, height = 10, units = "in", dpi = 300)
  #ggsave(paste0("plots/Microbial_Guilds/Fam_guild_", file_safe_g, ".tiff"),
         #plot = p_nolegend, width = 15, height = 10, units = "in", dpi = 600)

  # Save corresponding legend
  svglite::svglite(paste0("plots/Microbial_Guilds/Family_Level/Legends/Fam_legend_", file_safe_g, ".svg"),
                   width = 6, height = 6)
  grid::grid.draw(legend_grob)
  dev.off()

  tiff(paste0("plots/Microbial_Guilds/Family_Level/Legends/Fam_legend_", file_safe_g, ".tiff"),
       width = 6, height = 6, units = "in", res = 600)
  grid::grid.draw(legend_grob)
  dev.off()
}


```



