---
title: "Code 6: Statistical Analysis"
author: "Bryan Rodriguez"
date: "2025-03-11"
output: html_document
---

***Disclaimer: This R Markdown script was created by Bryan Rodriguez Colon with some use of AI (chatGPT-4o & chatGPT-5).***

## Overview

This R Markdown performs the core statistical analyses for the 16S amplicon dataset and generates summary visualizations used in the manuscript. It computes alpha diversity metrics and tests for differences among metadata-defined groups, and it evaluates beta diversity using Bray–Curtis dissimilarities with ordination (PCoA) and PERMANOVA to assess compositional differences among groups.

This document also includes a Venn diagram workflow that compares shared vs. unique taxa among lagoons based on presence in the relative-abundance tables grouped by `Sample_name`.


*NOTE: Optional blocks for UniFrac and Aitchison/CLR workflows are present but can be disabled if not used in the manuscript.*



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

setwd("/Users/bryanrodriguez/Library/CloudStorage/OneDrive-UniversityofKansas/Bryan-MS_PhD/PHD-PROJECT/PuertoRico/CHAPTER2/16S/Processing/Geobio_Guanica_Codes/R_Codes")

rm(list = ls())


```


```{r Load packages, echo = F, include=F}
library(ggplot2)    # Core plotting system used for alpha diversity boxplots, ordinations (PCoA), and saving figures
library(phyloseq)   # Handles microbiome data structures and provides diversity metrics, distance calculations, and ordination tools
library(vegan)      # Provides ecological statistics, including Bray–Curtis distances and PERMANOVA (adonis2)
library(tidyr)      # Reshapes data for statistical summaries and plotting (e.g., long-format alpha diversity tables)
library(colorspace) # Adjusts and standardizes color brightness and contrast for consistent figure aesthetics
library(svglite)    # Enables high-quality SVG output suitable for vector editing and publication
library(dplyr)      # Data manipulation framework used for filtering, grouping, summarizing, and table construction
library(pheatmap)   # Generates heatmaps for optional beta-diversity or distance-matrix visualizations
library(ggvenn)     # Draws Venn diagrams to visualize shared and unique taxa among lagoons
library(tibble)
```

# Loading and Preparing data files and variables

## Load Data

```{r Load phyloseq files, echo=F, include=FALSE}
#Non-nomalized data
load(file = "OutputData/RData/GuanicaGeobio_Arch_050525.RData")
load(file = "OutputData/RData/GuanicaGeobio_Bac_050525.RData")
load(file = "OutputData/RData/GuanicaGeobio_Arch_Bac_050525.RData")

#Normalized data
load(file = "OutputData/RData/GuanicaGeobio_Arch_Bac_norm_050525.RData")
load(file = "OutputData/RData/GuanicaGeobio_Bac_norm_050525.RData")
load(file = "OutputData/RData/GuanicaGeobio_Arch_norm_050525.RData")

#For Venn Diagram
melt_file_list <- readRDS("OutputData/RData/PE_GuanicaGeobio_Arch_Bac_merged_rel_melt_SLayer.RDS")

Arch_Bac_merged_rel_melt <- melt_file_list$Arch_Bac_merged_rel_melt_SLayer
Arch_merged_rel_melt <- melt_file_list$Arch_merged_rel_melt_SLayer
Bac_merged_rel_melt <- melt_file_list$Arch_Bac_merged_rel_melt_SLayer


```

## Variables for the statistical analysis

```{r}
# Define the desired sample order
sample_order_All <- c("POE1-gc", "POE1-g", "POE1-R", "POE1-B", 
                       "JB-C", "JB-g", "JB-DP", "JB-B", 
                      "PS-C1", "PS-C2", "PS-Br", "PS-L", "PS-Bottom")

# Ensure the Stromatolite_Layer column is a factor with the desired levels
Arch_Bac@sam_data$Sample_Layer <- factor(Arch_Bac@sam_data$Sample_Layer, levels = sample_order_All)


```



# Alpha Diversity

## Alpha Diversity Calculations

```{r Calculations}


# Calculate alpha diversity metrics (Observed, Shannon, Simpson)
alpha_div_Bac <- estimate_richness(Arch_Bac, measures = c("Observed", "Shannon", "Simpson"))
sample_metadata <- sample_data(Arch_Bac)

alpha_div_Bac$Sample <- rownames(alpha_div_Bac)
alpha_div_Bac <- merge(alpha_div_Bac, sample_metadata, by.x = "Sample", by.y = "row.names")


# Add a column for the main sample group (e.g., "PO_C3", "PO_E1", "JB", "PS")
# Modify the name of this column if needed
alpha_div_Bac$Sample_name <- factor(alpha_div_Bac$Sample_name, levels = c("POC3", "POE1", "JB", "PS"))


```

## Tables for Publication

```{r Table for paper}

# Summarize alpha diversity metrics by Sample_name
alpha_summary <- alpha_div_Bac %>%
  group_by(Sample_Layer) %>%
  summarise(
    Observed_mean = mean(Observed), Observed_sd = sd(Observed),
    Shannon_mean = mean(Shannon), Shannon_sd = sd(Shannon),
    Simpson_mean = mean(Simpson), Simpson_sd = sd(Simpson),
  )

# Round numeric columns only
alpha_summary[, -1] <- round(alpha_summary[, -1], 2)

# View table
print(alpha_summary)


# Optional: Save to CSV
write.csv(alpha_summary, "OutputData/Statistics/alpha_diversity_summary.csv", row.names = FALSE)



```


## Plotting

```{r Alpha Diversity Plot in Bacteria and Archaea, echo=F, include=F }

# Pivot data for long format to enable faceting
alpha_long_Bac <- alpha_div_Bac %>%
  pivot_longer(cols = c("Observed", "Shannon", "Simpson"), names_to = "Metric", values_to = "Value")


# Define colors
colors_To_use <- c("POC3" = "green", "POE1" = "blue", "JB" = "red", "PS" = "purple")

# Create less darker versions for the stroke
fill_color <- darken(colors_To_use, amount = 0.2)
Stroke_color <- darken(colors_To_use, amount = 0.5)

# Create the plot
alpha_plot_Bac <- ggplot(alpha_long_Bac, aes(x = Sample_Layer, y = Value, fill = Sample_name, color = Sample_name)) +
  geom_boxplot(outlier.size = 0.5) +
  scale_fill_manual(values = fill_color) +
  scale_color_manual(values = Stroke_color) +     # Apply stroke colors
  facet_wrap(~ Metric, scales = "free_y", ncol = 1) +
  labs(fill = "Sample", color = "Sample") +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1),
    plot.title = element_text(hjust = 0.5)
  ) +
  xlab("Microbial mat and microbialite samples layers") +
  ylab("Alpha Diversity") +
  ggtitle("Alpha Diversity Metrics (Shannon, Observed, Simpson, Chao1)")

# Display
#ggplotly(alpha_plot_Bac)

alpha_plot_Bac

ggsave(
  filename = "plots/Statistics/alpha_plot_Bac.tiff",       # File name with .tiff extension
  plot = alpha_plot_Bac,                 # The plot object to save
  width = 6,                             # Width in inches
  height = 10,                            # Height in inches
  dpi = 300                              # Resolution (300 dpi for high-quality print)
)


```


# Beta Diversity

## Bray-Curtis distance

```{r Bray-Curtis distance}

# Calculate Bray-Curtis distance
bray_dist_Bac <- phyloseq::distance(Arch_Bac_norm, method = "bray")

# Perform PCoA ordination
ordination_Bac <- ordinate(Arch_Bac, method = "PCoA", distance = bray_dist_Bac)

# Extract eigenvalues and % variance explained
eig_vals <- ordination_Bac$values$Eigenvalues
var_explained <- eig_vals / sum(eig_vals)
axis1_pct <- round(100 * var_explained[1], 1)
axis2_pct <- round(100 * var_explained[2], 1)

# Extract ordination data with metadata
ordination_df <- as.data.frame(ordination_Bac$vectors)
ordination_df$SampleID <- rownames(ordination_df)
ordination_df <- merge(
  ordination_df,
  as.data.frame(sample_data(Arch_Bac_norm)),
  by.x = "SampleID",
  by.y = "row.names"
)

# Define custom shape values (13 unique shapes)
shape_values <- c(
  "POE1-gc" = 0, "POE1-g" = 1, "POE1-R" = 2, "POE1-B" = 3,
  "JB-C" = 4, "JB-g" = 5, "JB-DP" = 6, "JB-B" = 7,
  "PS-C1" = 8, "PS-C2" = 9, "PS-Br" = 10, "PS-L" = 13, "PS-Bottom" = 12
)

# Create the plot
Beta_div_Bac_PCoA <- ggplot(ordination_df, aes(x = Axis.1, y = Axis.2)) +
  geom_point(aes(color = Sample_name, shape = Sample_Layer), size = 3) +
  stat_ellipse(aes(color = Sample_name), type = "norm", linetype = "dashed") +
  scale_shape_manual(values = shape_values) +
  ggtitle("PCoA Plot (Bray-Curtis)") +
  theme_classic() +
  theme(
    panel.grid = element_blank(),
    axis.line = element_line(color = "black"),
    legend.title = element_blank(),
    plot.title = element_text(hjust = 0.5)
  ) +
  labs(
    x = paste0("PCoA Axis 1 (", axis1_pct, "%)"),
    y = paste0("PCoA Axis 2 (", axis2_pct, "%)")
  )

Beta_div_Bac_PCoA
# Display the plot

# Save as SVG for Illustrator
ggsave(
  filename = "plots/Statistics/Beta_div_Bac_PCoA.svg",
  plot = Beta_div_Bac_PCoA,
  width = 8,
  height = 6
)

```

## PERMANOVA Analysis

```{r PERMANOVA, include=T, echo=T}
# Extract sample metadata for grouping
metadata_Bac <- as(sample_data(Arch_Bac_norm), "data.frame")

# Perform PERMANOVA
permanova_Bac <- adonis2(bray_dist_Bac ~ Sample_name, data = metadata_Bac, permutations = 999)

# Extract and format PERMANOVA results into a table
summary_table <- as.data.frame(permanova_Bac) %>%
  rownames_to_column(var = "Source") %>%
  mutate(
    Component = SumOfSqs / sum(SumOfSqs),   # Variance component
    Var = SumOfSqs / Df,                   # Variance
    SD = sqrt(Var)                         # Standard deviation
  )

# Rename columns for clarity
colnames(summary_table) <- c("Source", "Df", "SS", "MS", "Pseudo F", "p-value", "Component", "Var", "SD")

# View the table
print(summary_table)

# Save the table to a CSV for your paper
write.csv(summary_table, "OutputData/Statistics/PERMANOVA_summary_detailed_norm.csv", row.names = FALSE)

cat("Detailed PERMANOVA summary saved to 'PERMANOVA_summary_detailed_norm.csv'")
```

## Unifrac Analysis (NOT USED)


```{r unifrac analysis}

# Calculate Bray-Curtis distance
unifrac_dist <- phyloseq::distance(Arch_Bac_norm, method = "unifrac")

# Compute hierarchical clustering
hclust_unifrac <- hclust(as.dist(unifrac_dist), method = "average")  # UPGMA

# Plot dendrogram
plot(hclust_unifrac, main = "Hierarchical Clustering of Samples (UniFrac)", xlab = "", sub = "")



# Convert distance matrix to matrix
dist_matrix <- as.matrix(unifrac_dist)

# Plot heatmap
pheatmap(dist_matrix, clustering_method = "average", main = "UniFrac Distance Heatmap")


```

## Aitchison Analysis (NOT USED)


```{r aitchison analysis}
library(compositions)

# Extract OTU table from phyloseq
otu_mat <- as(otu_table(Arch_Bac_norm), "matrix")

# If taxa are rows and samples are columns, transpose
if (taxa_are_rows(Arch_Bac_norm)) {
  otu_mat <- t(otu_mat)
}

# Add pseudocount (e.g., 0.5)
otu_mat_nozeros <- otu_mat
otu_mat_nozeros[otu_mat_nozeros == 0] <- 0.5




# Perform CLR transformation
otu_clr <- clr(otu_mat_nozeros)


aitchison_dist <- dist(otu_clr, method = "euclidean")


# Classical multidimensional scaling
aitchison_pcoa <- cmdscale(aitchison_dist, k = 2, eig = TRUE)

# Extract coordinates
ordination_df <- as.data.frame(aitchison_pcoa$points)
colnames(ordination_df) <- c("Axis.1", "Axis.2")
ordination_df$SampleID <- rownames(ordination_df)

# Merge metadata
metadata_df <- data.frame(as(sample_data(Arch_Bac_norm), "data.frame"))
ordination_df <- merge(ordination_df, metadata_df, by.x = "SampleID", by.y = "row.names")




p_aitchison <- ggplot(ordination_df, aes(x = Axis.1, y = Axis.2, color = Sample_name)) +
  geom_point(size = 3) +
  stat_ellipse(type = "norm", linetype = "dashed") +
  ggtitle("PCoA on Aitchison Distance (CLR-Euclidean)") +
  theme_classic() +
  theme(
    legend.title = element_blank(),
    plot.title = element_text(hjust = 0.5)
  )

print(p_aitchison)


```


# Venn Diagram


## Splitting Data by sample
Splitting Data by Sample_name summarizing abundance data at each taxonomic level and sorting abundance in descending order. 
```{r split data by sample layer}
Arch_merged_rel_melt_group <- Arch_merged_rel_melt %>% dplyr::group_by(Sample_name, OTU, Phylum, Class, Order, Family, Genus) %>% dplyr::summarize(Rel_Abund = sum(RelAbund), Rel_Abund_SD = sd(RelAbund)) %>% arrange(desc(Rel_Abund))

Bac_merged_rel_melt_group <- Bac_merged_rel_melt %>% dplyr::group_by(Sample_name, OTU, Phylum, Class, Order, Family, Genus) %>% dplyr::summarize(Rel_Abund = sum(RelAbund), Rel_Abund_SD = sd(RelAbund)) %>% arrange(desc(Rel_Abund))

Arch_Bac_merged_rel_melt_group <- Arch_Bac_merged_rel_melt %>% dplyr::group_by(Sample_name, OTU, Phylum, Class, Order, Family, Genus) %>% dplyr::summarize(Rel_Abund = sum(RelAbund), Rel_Abund_SD = sd(RelAbund)) %>% arrange(desc(Rel_Abund))

rm(Arch_merged_rel_melt, Bac_merged_rel_melt, Arch_Bac_merged_rel_melt)

```



## Filter by name


```{r Filter by name}
# Filter by lagoon and extract unique Genus
OTU_PO <- Arch_Bac_merged_rel_melt_group %>%
  filter(Sample_name == "POE1", Rel_Abund > 0, !is.na(OTU)) %>%
  pull(OTU) %>%
  unique()

OTU_JB <- Arch_Bac_merged_rel_melt_group %>%
  filter(Sample_name == "JB", Rel_Abund > 0, !is.na(OTU)) %>%
  pull(OTU) %>%
  unique()

OTU_PS <- Arch_Bac_merged_rel_melt_group %>%
  filter(Sample_name == "PS", Rel_Abund > 0, !is.na(OTU)) %>%
  pull(OTU) %>%
  unique()


```

## Plot Venn Diagram

```{r Plot}

Genus_list <- list(PO = OTU_PO, JB = OTU_JB, PS = OTU_PS)

venn_plot <- ggvenn(Genus_list, 
       fill_color = c("blue", "red", "green"), 
       text_size = 3, 
       set_name_size = 4)


ggsave(
  filename = "plots/Statistics/venn_plot_PerSample.svg",
  plot = venn_plot,
  width = 6,
  height = 6
)

venn_plot


```






